---
title: "[Insert title here]"
author: "Timothy"
date: "`r Sys.Date()`"

abstract: "[Insert abstract here]"

output:
  html_document:
    toc: true
    toc_float: true
    theme: cerulean
    code_folding: show
  editor_options:
    chunk_output_type: inline
---

# Motivation


## Background



# Settings
Chunk options were set.
```{r}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  error     = FALSE,
  fig.width = 14,
  fig.height = 7,
  out.width = "100%")
```


# Setup


## Environment
Environment was cleaned, and outputs were cleared. The seed is set here.
```{r}
rm(list = ls(all.names = TRUE)) # remove objects
cat("\014")                     # clear
seed <- 4245
options(scipen = 999)
```


## Packages
Libraries were loaded.
```{r}
packages <- c(
  "conflicted",
  "lmtest",
  "sjPlot",
  "vip",
  "rpart",
  "rpart.plot",
  "car",
  "readxl",
  "smotefamily",
  "fastshap",
  "shapviz",
  "hstats",
  "pdp",
  "gbm",
  "rstatix",
  "pROC",
  "ggcorrplot",
  "missRanger",
  "glmnet",
  "openxlsx",
  "gridExtra",
  "psych",
  "Hmisc",
  "scales",
  "reshape2",
  "purrr",
  "tibble",
  "rmarkdown",
  "knitr",
  "tidyverse",
  "caret",
  "dplyr")

invisible(lapply(
  packages,
  \(p) {
    if (!requireNamespace(p, quietly = TRUE)) install.packages(p)
    library(p, character.only = TRUE)}))
```


## Functions
Functions were deconflicted with conflict_prefer() from the **conflicted** package (Wickham, 2024).
```{r, message = FALSE, warning = FALSE}
conflict_scout()

functions <- c(
  filter = "dplyr",
  lag = "dplyr",
  combine = "dplyr",
  recode = "dplyr",
  select = "dplyr",
  alpha = "ggplot2",
  partial = "pdp",
  describe = "psych",
  summarize = "dplyr",
  cbind = "base",
  t_test = "rstatix")

invisible(lapply(names(functions), function(f) conflict_prefer(f, functions[f])))
```


# Data Cleaning


## Dataset
The raw dataset is accessible at: https://osf.io/7xcmy/files/f68nh.

The codebook is accessible at: https://www.icpsr.umich.edu/web/NAHDAP/series/706/variables.


## Import
The raw dataset was imported with read_excel() from the **readxl** package (Wickham & Bryan, 2025).
```{r, message = FALSE, warning = FALSE}
df <- read_excel(
  "File 4 - NSHAP WAVE 2 - Education and cancer.xlsx",
  sheet = "34921-0001-Data",
  skip = 10,
  col_names = TRUE)
```


## Variables
Variables were selected.
```{r}
df <- df %>%
  select(dia = MOCA...17                       # outcome
       , alc = DRNKWKLY                        # moderator
       , job = HRSCJOB
       , smk = AVECIG
       , phy = PHYSACT2...12
       , job = HRSCJOB
       , soc = `SOCIALLY ACTIVE`               # main effect
       
       , sex = GENDER...44
       , age = AGE                             # main effect
       , mar = `NEVER MARRIED...16`
       , edu = DEGREE_RECODE
       , eth = RACE_RECODE...45
       , inc = INCOME_1...930)

df %>% describe() %>% paged_table()
```


## Observations
Observations with missing or invalid values in the outcome (**dia**) were excluded case-wise. 

Dementia status was classified according to established clinical cutoffs from the Montreal Cognitive Assessment (MoCA) scale (Carson et al., 2018).
```{r}
df <- df %>%
  mutate(
    dia = case_when(
      dia > 0 & dia < 18.5 ~ "1",
      dia >= 18.5 ~ "0",
      TRUE ~ NA_character_)) %>%
  filter(!is.na(dia), dia %in% c("0", "1"))

df$dia %>% table()
df %>% dim()
```


## Values
Values were cleaned based on dataset documentation in the codebook.
```{r}
df$smk[df$smk == -3] <- 0
df$job[df$job == -3] <- 0
df$alc[df$alc == -3] <- 0

na_rules <- list(mar = list(discrete = c(), range = list(c(0, 1)))
               , smk = list(discrete = c(), range = list(c(0, Inf)))
               , phy = list(discrete = c(-3), range = list(c(1, 5)))
               , sex = list(discrete = c(0, -3, -2, -1), range = list(c(1, 2)))
               , edu = list(discrete = c(0, -3, -2, -1, 7), range = list(c(1, 6)))
               , eth = list(discrete = c(0, -3, -2, -1), range = list(c(1, 3)))
               , inc = list(discrete = c(-5, -4, -2, -1), range = list(c(1, 5)))
               , age = list(discrete = c(0, -8, -6, -5, -4, -3, -2, -1), range = list(c(61, Inf)))
               , alc = list(discrete = c(-8, -3, -2, -1), range = list(c(0, 7)))
               , job = list(discrete = c(-8, -6, -5, -4, -3, -2, -1), range = list(c(0, Inf)))
               , soc = list(discrete = c(0), range = list(c(6, 32)))
               , dia = list(discrete = c(-1), range = list(c(0, 1))))

vars <- names(na_rules)

convert_to_na <- function(x, discrete_vals, ranges) {
  is_na <- x %in% discrete_vals
  for (r in ranges) is_na <- is_na | (x < r[1] | x > r[2])
  x[is_na] <- NA
  x}

df[vars] <- lapply(vars, \(v) convert_to_na(df[[v]], na_rules[[v]]$discrete, na_rules[[v]]$range))

df %>% describe() %>% paged_table()
```


## Missing
Missing values were identified and tallied.
```{r}
colSums(is.na(df))

inval0 <- c("age") # a value of 0 would not be valid in these variables

sapply(df[inval0], function(x) sum(x == 0, na.rm = TRUE))

missing <- round(sum(is.na(df)) / prod(dim(df)) * 100, 2)

print(paste("Percentage of data missing:", missing, "%"))
```


## Factors
Categorical variables were factorised and reference categories were defined.
```{r}
df$sex <- factor(recode(as.character(df$sex), 
                        `1` = "Male", 
                        `2` = "Female")) %>% relevel(ref = "Male")

df$eth <- factor(recode(as.character(df$eth), 
                        `1` = "White", 
                        `1.29` = "Non-White",
                        `2` = "Non-White",
                        `3` = "Non-White")) %>% relevel(ref = "White")

df$mar <- factor(recode(as.character(df$mar), 
                       `0` = "Married", 
                       `1` = "NeverMarried")) %>% relevel(ref = "Married")

df$dia <- factor(recode(as.character(df$dia),
                        `0` = "Nondemented", 
                        `1` = "Demented")) %>% relevel(ref = "Nondemented")

df %>% describe() %>% paged_table()
```


# Data Exploration


## Variables
Variables were sorted into continuous, ordinal, and categorical variables. 
```{r}
group <- "dia"
ordvars <- c("phy", "edu", "inc")
numvars <- setdiff(names(df)[sapply(df, is.numeric)], c(group, ordvars))
catvars <- setdiff(names(df)[sapply(df, is.factor) & !sapply(df, is.ordered)], group)
```


## Distributions
Distributions of each predictor by dementia status are visualised.

No obvious class separation was observed for any variable.

The number of cigarettes per day (**smk**), number of days per week alcohol is consumed (**alc**), and job hours per week (**job**) were transformed into nominal categories representing presence or absence. 
```{r}
df_num_long <- df %>%
  select(all_of(c(group, numvars))) %>%
  pivot_longer(cols = all_of(numvars), names_to = "variable", values_to = "value") %>%
  filter(!is.na(value))

for (v in unique(df_num_long$variable)) {
  dfnona <- subset(df_num_long, variable == v & !is.na(value))
  p <- ggplot(dfnona, aes(x = value, fill = .data[[group]])) +
    geom_density(alpha = 0.4) +
    theme_minimal() +
    labs(title = paste("Density of", v), subtitle = paste("Grouped by", group), x = v, y = "Density")
  print(p)}

for (v in ordvars) {
  dfnona <- df %>%
    filter(!is.na(.data[[v]]), !is.na(.data[[group]])) %>%
    mutate(ord_var = factor(.data[[v]], levels = sort(unique(.data[[v]])), ordered = TRUE))
  p <- ggplot(dfnona, aes(x = ord_var, fill = .data[[group]])) +
    geom_bar(position = "dodge", color = "black", width = 0.6) +
    theme_minimal() +
    labs(title = paste("Count of", group, "by", v), x = v, y = "Count")
  print(p)}

for (v in catvars) {
  dfnona <- df %>%
    filter(!is.na(.data[[v]]), !is.na(.data[[group]])) %>%
    group_by(across(all_of(c(v, group)))) %>%
    summarize(n = n(), .groups = "drop") %>%
    group_by(across(all_of(v))) %>%
    mutate(prop = n / sum(n)) %>%
    ungroup()
  p <- ggplot(dfnona, aes(x = .data[[v]], y = prop, fill = .data[[group]])) +
    geom_col(position = "fill", color = "black", width = 0.3) +
    theme_minimal() +
    labs(title = paste("Proportion of", group, "by", v), x = v, y = "Proportion")
  print(p)}
```


## Discretisation
The distributions of the transformed variables are shown here.

The characteristics of the overall sample, and subsets split by dementia status, are displayed.
```{r}
df$job <- factor(recode(as.character(ifelse(df$job == 0, 0, 1)), `0` = "No.Job", `1` = "Have.Job")) %>% relevel(ref = "No.Job")
df$smk <- factor(recode(as.character(ifelse(df$smk == 0, 0, 1)), `0` = "No.Smoke", `1` = "Have.Smoke")) %>% relevel(ref = "No.Smoke")
df$alc <- factor(recode(as.character(ifelse(df$alc == 0, 0, 1)), `0` = "No.Alcohol", `1` = "Have.Alcohol")) %>% relevel(ref = "No.Alcohol")

numvars <- setdiff(names(df)[sapply(df, is.numeric)], group)
numvars <- setdiff(numvars, ordvars)
catvars <- setdiff(names(df)[sapply(df, is.factor) & !sapply(df, is.ordered)], group)

for (v in catvars) {
  dfnona <- df %>%
    filter(!is.na(.data[[v]]), !is.na(.data[[group]])) %>%
    group_by(across(all_of(c(v, group)))) %>%
    summarize(n = n(), .groups = "drop") %>%
    group_by(across(all_of(v))) %>%
    mutate(prop = n / sum(n)) %>%
    ungroup()
  
  p <- ggplot(dfnona, aes(x = .data[[v]], y = prop, fill = .data[[group]])) +
    geom_col(position = "fill", color = "black", width = 0.3) +
    theme_minimal() +
    labs(title = paste("Proportion of", group, "by", v), x = v, y = "Proportion")
  
  print(p)}

df %>% describe() %>% paged_table()
df %>% filter(dia == "Nondemented") %>% describe() %>% paged_table()
df %>% filter(dia == "Demented") %>% describe() %>% paged_table()

table(df$dia, df$sex)
table(df$dia, df$eth)
table(df$dia, df$mar)
table(df$dia, df$alc)
table(df$dia, df$job)
table(df$dia, df$smk)
```


## Differences
Differences by dementia status were assessed. 

Levene's tests, independent-samples t-tests, and Wilcoxon tests were run using the **rstatix** package (Kassambara, 2023). Chi-square tests for categorical variables were run.

*If the p-values are not visible in the table, use the navigation arrows in the top-right of each paged table.*
```{r}
levene_results <- map_dfr(numvars, ~ {
  df %>%
    filter(!is.na(.data[[.x]]), !is.na(.data[[group]])) %>%
    levene_test(as.formula(paste(.x, "~", group))) %>%
    mutate(variable = .x)})

unequalnumvars <- levene_results %>%
  filter(p < 0.05) %>%
  pull(variable) # var.equal = FALSE

equalnumvars <- setdiff(numvars, unequalnumvars) # var.equal = TRUE

ttest_equal_results <- df %>%
  pivot_longer(
    cols = all_of(equalnumvars),
    names_to = "variable",
    values_to = "value") %>%
  filter(!is.na(value), !is.na(.data[[group]])) %>%
  group_by(variable) %>%
  t_test(as.formula(paste("value ~", group)), var.equal = TRUE) %>%
  ungroup()

ttest_unequal_results <- map_dfr(unequalnumvars, ~ {
  df %>%
    filter(!is.na(.data[[.x]]), !is.na(.data[[group]])) %>%
    t_test(as.formula(paste(.x, "~", group)), var.equal = FALSE) %>%
    mutate(variable = .x)})

wilcox_results <- df %>%
  pivot_longer(
    cols = all_of(ordvars),
    names_to = "variable",
    values_to = "value") %>%
  filter(!is.na(value), !is.na(.data[[group]])) %>%
  group_by(variable) %>%
  wilcox_test(as.formula(paste("value ~", group)), exact = FALSE) %>%
  ungroup()

chisq_results <- map_dfr(catvars, ~ {
  tbl <- table(df[[.x]], df[[group]])
  res <- chisq.test(tbl)
  broom::tidy(res) %>%
    mutate(variable = .x)})

levene_results %>% select(variable, p) %>% paged_table()
ttest_equal_results %>% select(variable, p) %>% paged_table()
ttest_unequal_results %>% select(variable, p) %>% paged_table()
wilcox_results %>% select(variable, p) %>% paged_table()
chisq_results %>% select(variable, p.value) %>% rename(p = p.value) %>% paged_table()
```


## Correlations
Spearman correlations between continuous and ordinal predictors were calculated with rcorr() from the **Hmisc** package (Harrell, 2019).
```{r}
corrvars <- c(ordvars, numvars)
df_num <- df[, corrvars]
cor_res <- rcorr(as.matrix(df_num), type = "spearman")
cor_matrix <- cor_res$r
p_matrix <- cor_res$P

ggcorrplot(cor_matrix, 
           hc.order = TRUE, 
           type = "upper",
           lab = TRUE, 
           p.mat = p_matrix, 
           sig.level = 0.05,
           insig = "blank",
           colors = c("violetred", "white", "steelblue"))
```


## Multicollinearity
Variance inflation factors were computed with vif() from the **car** package (Fox & Weisberg, 2019). 
```{r}
vif(lm(as.numeric(dia) ~ ., data = df))
```


